% ── encoding.tex · §2 Preliminaries + §3 Encoding ──────────────────
\section{Preliminaries}\label{sec:prelim}

\subsection{The Predicate Prover}\label{sec:pp-syntax}

The syntax for the \emph{predicates} and \emph{expressions} of \pp{} is
given as follows:
\begin{align*}
	E, F & \Coloneqq x \mid n \mid E \mapsto F \mid F(E_1,\ldots,E_k)
	\mid E + F \mid E - F \mid -E                                     \\
	P, Q & \Coloneqq E \mid E = F \mid E \leq F
	\mid E_1,\ldots,E_k \in F
	\mid \lnot P \mid P \land Q \mid P \lor Q \mid P \Rightarrow Q
	\mid P \Leftrightarrow Q
	\mid \mathcal{Q}\,\vec{x} \cdot P
\end{align*}
where $x$ ranges over identifiers, $n$ over numerals, $\mapsto$ is B's pairing,
function application $F(E_1,\ldots,E_k)$ applies an expression $F$ to the arguments $E_1,\ldots,E_k$,
membership relates a tuple of expressions to an expression.
%
An expression may itself stand as an atomic predicate, shown by the leading
$E$ in the production for $P$.
%
A quantifier $\mathcal{Q} \in \{\forall, \exists, \Dia, \Hrt\}$ binds a
list of variables; $\forall (x,y) \cdot P$ is distinct from $\forall x \cdot \forall y
	\cdot P$, and dedicated rules (\rn{ALL1--4}, \rn{XST1--4}) regroup the
latter into the former during proof search.
%
The quantifiers $\Dia$ and $\Hrt$ are universal quantifiers idiosyncratic
to \pp's normalisation procedure; semantically both are $\forall$ and differ
only in the role they play during proof search.

Although B is a set-theoretic formalism, \pp{} itself proves formulas of
first-order logic with linear arithmetic: a pre-processing step reduces much
of the set theory of a proof obligation to first-order formulas over
membership before \pp{} sees it.
%
The set-theoretic operators that survive this reduction --- intervals, the
maplet, and the relational operators of \autoref{app:operators} --- reach
\pp{} as uninterpreted function symbols: \pp{} has no set-theory rules, and
such terms are only ever copied, compared for equality, or instantiated.

\pp{} proves \emph{sequents} $H \seq P$, where $H$ is a list of
hypothesis predicates and $P$ a goal predicate.
% 130 97
Alongside sequents, its proofs manipulate a second judgment, the
\emph{result}: we write $(H \seq P) \yields [Q_1, \ldots, Q_n]$ for
``under $H$, $P$ normalises to the list of predicates
$[Q_1, \ldots, Q_n]$'', whose conjunction --- taken only when the list is
consumed, by \rn{ALL7}/\rn{XST8} below --- is the normal form of $P$.
%
\pp's inference rules are specified in the annex of its internal
documentation~\cite{pp_spec}: roughly \numRulesSpec{} rules, grouped by
families — propositional (\rn{AND}, \rn{OR}, \rn{IMP}, \rn{EQV},
\rn{NOT}), hypothesis discharge (\rn{AXM1--9}), quantifiers
(\rn{ALL1--9}, \rn{XST1--8}), equality
(\rn{EQL}, \rn{EAXM}, \rn{EQC}, \rn{ECTR}, \ldots), booleans
(\rn{BOOL}), linear arithmetic (\rn{AR1--13}),
and normalisation (\rn{NRM1--30}).
%
Each rule reduces the current goal to zero, one, or two
premises, possibly under computational side-conditions
(``$P$ occurs in the list of hypotheses $H$'',
``$a$ and $b$ are integer literals such that $a + b < 0$'', etc.).
%
A rule whose conclusion is a \emph{result} computes its list by one of
three schemas: the empty list (the formula is discharged), a singleton
$[Q]$ (a single transformed formula), or the concatenation of its two
premises' lists (a combination).
%%GB: Shall we say that to each "classical" rule corresponds a "result" rule, computed as follow except some special cases?
%
For instance, \rn{AND4} introduces a conjunct, and its result form
\rn{AND4\_1} (the combination schema) concatenates the two premises'
lists:
\begin{center}
  \begin{prooftree}
    \hypo{H \seq Q} \hypo{H \seq P}
    \infer2[\rn{AND4}]{H \seq P \land Q}
  \end{prooftree}
  \qquad
  \begin{prooftree}
    \hypo{(H \seq Q) \yields R_1} \hypo{(H \seq P) \yields R_2}
    \infer2[\rn{AND4\_1}]{(H \seq P \land Q) \yields R_1 \app R_2}
  \end{prooftree}
\end{center}
\rn{ALL7} and \rn{XST8} are the only rules that \emph{consume} a result,
each turning a result derivation back into an ordinary sequent subproof.
%
\rn{ALL7} normalises the body of a universal antecedent pointwise to a list
$R_1,\ldots,R_n$ and continues under its conjunction, beneath the
post-normalisation universal $\Dia$:
\begin{center}
  \begin{prooftree}
    \hypo{(H \seq P) \yields [R_1,\ldots,R_n]}
    \hypo{H \seq (\Dia x \cdot R_1 ∧ \ldots ∧ R_n) \Rightarrow G}
    \infer2[\rn{ALL7}]{H \seq (\forall x \cdot P) \Rightarrow G}
  \end{prooftree}
\end{center}
%
\rn{XST8} is its existential dual: to prove $\exists x \cdot P$ it normalises
$\lnot P$ to a list $R_1,\ldots,R_n$ and continues from a refutation of the
universally-closed conjunction.
\begin{center}
  \begin{prooftree}
    \hypo{(H \seq \lnot P) \yields [R_1,\ldots,R_n]}
    \hypo{H \seq (\forall x \cdot R_1 ∧ \ldots ∧ R_n) \Rightarrow \bot}
    \infer2[\rn{XST8}]{H \seq \exists x \cdot P}
  \end{prooftree}
\end{center}
%
We encode the result judgment as a dedicated \lp{} type constructor, and the
rules that produce and consume it as functions over that type, in
\autoref{sec:enc-rules}.

As discussed in \autoref{sec:intro}, \pp{} is instrumented to record a
\trace{} file of the rules it applied; a companion utility re-runs the
proof and expands the trace into a \replay{} file that annotates every
step with the goal it was applied to.
%
Each replay line carries a rule name and possibly an argument
that follows the name in parentheses.
%
An example replay and its proof tree are shown in \autoref{fig:replay}.
%
The main proof is recorded prefix-style (a rule precedes its
subproofs), with two irregularities: the result derivation
feeding a rule with a result premise (i.e., \rn{ALL7}, \rn{XST8}) is
emitted \emph{before} that rule itself, and normalisation markers are
interleaved throughout.
%
The markers \texttt{NRM}, \texttt{FIN}, and \texttt{STOP\_NORM} bracket
\pp's normalisation passes --- \texttt{FIN} carries the predicate a pass
produced --- but none corresponds to an inference rule; they carry no proof
content, so \tool{} filters them out while rebuilding the tree.
%
The annotation of the first non-marker line is the overall goal.
%

% --- replay/derivation figure ----------------------------------------
\input{fig-replay.tex}
% ---------------------------------------------------------------------

\subsection{\lp}\label{sec:lp}

\lp~\cite{hondet:hal-02981561} is a proof assistant for
Cousineau and Dowek's \lpcmr~\cite{CousineauDowek},
the calculus underlying the \dedukti{} logical framework~\cite{assaf2023dedukti}.
%
The term syntax is given as follows:
\[
	t, u \Coloneqq \TYPE \mid \KIND \mid x \mid c
	\mid t\,u \mid \lambda (x : t),\, u \mid \Pi (x : t),\, u
\]
where $x$ ranges over variables, $c$ ranges over declared constants,
and terms are identified up to $\alpha$-equivalence.
%
Typing is standard for the \lpcmr{} (the full system is in
\autoref{app:typing}); the rule we rely on throughout applies types up to
\emph{conversion}: a term of type $A$ also has type $B$ whenever
$A \Equiv B$, where the conversion relation $\Equiv$ is generated by
$\beta$-equivalence together with the user-declared rewrite rules.
%
A \lp{} development is a sequence of declarations.
%
The \lpkw{symbol} command introduces a typed constant, with explicit
$(x : A)$ and implicit $[x : A]$ parameters and an optional definition;
\lpkw{rule} $\ell ↪ r$ adds a rewrite rule over pattern variables
\texttt{\$x}, checked for type-preservation at declaration time;
\lpkw{inductive} generates an inductive type with constructors and a recursor.
%
Modifiers refine a symbol's status: \lpmod{constant} symbols admit no
rewrite rule at their head, \lpmod{injective} symbols may be inverted
during unification, and the definition of an \lpmod{opaque} symbol is
sealed once type-checked.
%
A definition may be given as a proof script (\lpkw{begin} \ldots{}
\lpkw{end}) whose tactics (e.g., \lpkw{assume}, \lpkw{refine},
\lpkw{rewrite}, \lpkw{have}, \lpkw{induction}) ultimately
elaborate to a proof term that the kernel re-checks.

We build on \lp's standard library~\cite{Blanqui2021}: object-level types are
encoded in a Tarski-style universe $\Setu$ with $\El : \Setu \to \TYPE$
mapping each set code to its type of inhabitants, and a type $\Prop$ of
propositions with $\Prf : \Prop \to \TYPE$ mapping each proposition
to its type of proofs.
%
Unlike \dedukti{} --- a bare type-checker for the same calculus --- \lp{}
adds implicit arguments $[x:A]$ and an interactive tactic language on top
of the same kernel; both are elaborated away, so a type-checked \lp{}
development reduces to a \dedukti{}-checkable proof term.

\section{Encoding \pp{} in \lp}\label{sec:encoding}

\subsection{Encoding \pp's Predicates and Expressions}\label{sec:enc-formulas}

The encoding is shallow, and we assume the library's classical axioms
for excluded middle and propositional extensionality.
%
All B expressions inhabit a single domain: a set code $\iota$, so an
expression is a term of type $\El\,\iota$ and a predicate is a term of
type $\Prop$.
%
The library's code $o : \Setu$, with $\El\,o ↪ \Prop$, lets predicates
also occur where `set' elements are expected; the conjunction lists below
rely on this.
%
% sentence is too long. rework. maybe a figure for type signatures and rewrite rules.
The kernel file \texttt{B.lp} declares the domain constants,
$\mapsto$ (pairing), $\epsilon$ (membership), $\mathsf{BOOL}$ with
$\mathsf{BTRUE}$ and $\mathsf{BFALSE}$, B's functional application, and
the arithmetic signature $\mathbf{0}, \mathbf{1}, +, -, \leq$, together
with their axioms: injectivity and congruence of pairing, set
extensionality, the two-element axioms for $\mathsf{BOOL}$, and for
arithmetic the theory of a discretely ordered abelian group
($+$ commutative and associative with identity $\mathbf{0}$ and inverse
$-$; $\leq$ a partial order, invariant under translation; and the
discreteness axioms $\mathbf{0} \leq \mathbf{1}$,
$\lnot(\mathbf{1} \leq \mathbf{0})$ and
$\lnot(x \leq \mathbf{0}) \Rightarrow \mathbf{1} \leq x$).
%
This is just enough to \emph{derive} \pp's arithmetic rules;
it is not a set-theoretic formalisation of $\mathbb{Z}$
(e.g., as equivalence classes of pairs of natural numbers).
%
Numerals are decimal literals of the library's $\mathbb{N}$, coerced
into the domain by an embedding that \lp{} inserts automatically.
%
Every other operator --- the set-theoretic operators of
\autoref{sec:pp-syntax} included --- is an uninterpreted constant.

\paragraph{List-based Conjunction.}
Most of \pp{}'s rules treat $\land$ as left-associative, so pattern
matching $p \land q \land r$ gives $(p \land q)$ as the first conjunct
and $r$ as the second.
%
However, some rules peel the leftmost conjunct, and some
act on conjuncts from arbitrary positions.
%
For the sake of flexibility, we use a list-based encoding of conjunction:
an inductive type $\mathbb{L}^{*}$ with empty list $∎$ and right append $∷$,
and an injective operator $\bwand : \mathbb{L}^{*}\,o \to \Prop$ with the
rewrite rules
$\bwand(∎ ∷ \texttt{\$p}) ↪ \texttt{\$p}$ and
$\bwand(\texttt{\$ps} ∷ \top) ↪ \bwand{\texttt{\$ps}}$.
%
It is crucial that this conjunction stays \emph{abstract}: the connection to
standard binary conjunction is not given by rewrite rules but axiomatised
separately, and used only in the soundness proofs for \pp's inference rules.
%
The list form then surfaces directly in a rule's type.  \pp's \rn{AND2}
refutes a conjunction $\bwand(ps ∷ p)$ by deriving $\lnot p$ from the
remaining conjuncts $\bwand ps$; its rule and its encoding
(\texttt{lp/rules/Conj.lp}) read
\begin{center}
  \begin{prooftree}
    \hypo{H \seq \bwand ps \Rightarrow \lnot p}
    \infer1[\rn{AND2}]{H \seq \lnot\,\bwand (ps ∷ p)}
  \end{prooftree}
  \qquad
  $\rn{AND2} : \Prf\,(\bwand ps \Rightarrow \lnot p)
             \to \Prf\,(\lnot\,\bwand (ps ∷ p)).$
\end{center}

\paragraph{Compound Binders.}
\pp{} distinguishes the compound binder $\forall (x,y) \cdot P$ from the
nested $\forall x \cdot \forall y \cdot P$, and its quantifier rules manipulate
a compound binder as a unit.
%
We mirror this with an inductive dependent type $\Tuple : ℕ → \TYPE$
with projections $\prj\,[n] : \Tuple\,n → ℕ → \El\,\iota$, and quantifiers
$\allI, \xstI : (\Tuple\,n \to \Prop) \to \Prop$
whose introduction and elimination principles are axiomatised bridges to
the $\Pi$-types of \lp{} (e.g.\
$(\Pi v{:}\,\Tuple\,n,\ \Prf\,(P\,v)) \to \Prf\,(\allI v, P\,v)$ and its
converse); $\Dia$ and $\Hrt$ are defined aliases of $\allI$.
%
These bridges are axioms rather than rewrite rules: a rule unfolding $\allI$
to a nested $\Pi$ would erase the quantifier from goals before any rule lemma
could match it, and recovering the predicate $P$ at each use would demand the
higher-order unification \lp{} does not attempt.
%
% we probably dont need the following text but its useful to have:
% -----------------------------------------------------------------
% The regrouping rules (\rn{ALL1--4}, \rn{XST1--4}) need no \lp{}
% counterpart: \tool{} regroups binders itself while building the goal, so
% the \lp{} goal always carries one compound binder and these rules are
% consumed without emitting anything (\rn{ALL2} in \autoref{fig:replay} is
% an instance).
%
% we probably don't need to make this point but it's useful to have:
% -------------------------------------------------------------------
% No rewrite rule unfolds $\allI$ to nested $\forall$: if the quantifier
% reduced away, the surface form would vanish from goals before any rule
% lemma could address it, and inferring the
% implicit predicate $P$ at rule applications would require inverting the
% reduction, a higher-order unification problem \lp{} does not attempt.
%

\subsection{Encoding Rules}\label{sec:enc-rules}
%
Each \pp{} rule is a \lp{} symbol whose type states the rule and whose
body \emph{proves} it.
%
The encoding of a rule's statement is positional: the conclusion sequent
$H \seq P$ appears as $\Prf\,P$; a premise that extends the hypotheses,
$H, P \seq Q$, appears as the function type $\Prf\,P \to \Prf\,Q$; and a
side-condition ``$P$ occurs in $H$'' becomes an explicit argument of
type $\Prf\,P$ (supplied by \tool{} at translation time, \autoref{sec:eval}).
%
For example, \rn{IMP4} (``to prove $P \Rightarrow Q$, assume $P$ and
prove $Q$'') and the discharge rule \rn{AXM1} (``$\lnot P$ occurs in
$H$, so $P \Rightarrow Q$ holds'') have the following type signatures:
\[
	\rn{IMP4} : (\Prf\,P \to \Prf\,Q) \to \Prf\,(P \Rightarrow Q)
	\qquad
	\rn{AXM1} : \Prf\,(\lnot P) \to \Prf\,(P \Rightarrow Q).
\]


\paragraph{Result Derivations.}
A result derivation (\autoref{sec:pp-syntax}) produces a list of predicates
standing for the normal form of a formula.  Write $\mathsf{conj}\,ts$ for the
conjunction of a snoc list $ts$ --- $\top$ when $ts$ is empty, and the
$\bwand$ of its elements otherwise.  We encode the derivation as a type
$\Res\,t$ whose value pairs the list with a proof that $t$ equals that
conjunction:
\begin{align*}
  \Res &: \El\,o \to \TYPE \\
  \mathsf{mk\_res}\,[t : \El\,o]\,(ts : \mathbb{L}^{*}\,o) &: \Prf\,(t = \mathsf{conj}\,ts) \to \Res\,t \\
  \mathsf{res\_ts}\,(\mathsf{mk\_res}\,ts\,h) &↪ ts \\
  \mathsf{res\_eq}\,(\mathsf{mk\_res}\,ts\,h) &↪ h \\
  \restm\,r &≔ \mathsf{conj}\,(\mathsf{res\_ts}\,r)
\end{align*}
\noindent A result-form rule (suffixed \texttt{\_1} in replays) is a
\lp{} function over $\Res$ values: \rn{AND4\_1} concatenates its two
children's lists, and \rn{ALL7} takes a per-instance result derivation
$\rho$ and continues under the conjunction ($\restm$) of its result.
\begin{codeLPmulti}
\begin{lstlisting}[style=lp]
symbol AND4_1 [ps : $\mathbb{L}^{*}$ o] [p : Prop]
    (r1 : Res p) (r2 : Res ($\bigwedge$ ps)) : Res ($\bigwedge$ (ps ∷ p))
  ≔ mk_res (concat (res_ts r1) (res_ts r2)) (...);
\end{lstlisting}
\listsep
\begin{lstlisting}[style=lp]
opaque symbol ALL7 [n] [P : Tuple n → Prop] [Q : Prop]
    (ρ : Π v : Tuple n, Res (P v))
    : π ((`$\diamondsuit$ v, res_tm (ρ v)) ⇒ Q) → π ((`!! v, P v) ⇒ Q) ≔ ...
\end{lstlisting}
\end{codeLPmulti}
\noindent For example, consider the result derivation below, for the body
$P_1 \land P_2 \land P_3$ with each conjunct kept at a \rn{STOP\_1} leaf:
\begin{center}\footnotesize
\ebproofset{separation=1em}%
\begin{prooftree}
  \infer0[\rn{STOP\_1}]{(\seq P_3) \yields [P_3]}
  \infer0[\rn{STOP\_1}]{(\seq P_2) \yields [P_2]}
  \infer0[\rn{STOP\_1}]{(\seq P_1) \yields [P_1]}
  \infer2[\rn{AND4\_1}]{(\seq P_1 \land P_2) \yields [P_2, P_1]}
  \infer2[\rn{AND4\_1}]{(\seq P_1 \land P_2 \land P_3) \yields [P_3, P_2, P_1]}
  \hypo{}
  \ellipsis{}{\seq (\Dia x \cdot P_3 \land P_2 \land P_1) \Rightarrow G}
  \infer2[\rn{ALL7}]{\seq (\forall x \cdot P_1 \land P_2 \land P_3) \Rightarrow G}
\end{prooftree}
\end{center}
The example demonstrates why results are lists rather than conjunctions:
the two nested \rn{AND4\_1} steps concatenate
their children's lists into the single flat list $[P_3, P_2, P_1]$, which
\rn{ALL7} conjoins.  Combining the results eagerly with $\land$ would
instead mirror the proof's nesting as $P_3 \land (P_2 \land P_1)$; the list
defers and flattens the conjunction.
%
Reconstructed, this derivation and its \rn{ALL7} step become the \lp{} proof
term
{\small
\[
  \rn{ALL7}\ (\lambda v,\ \rn{AND4\_1}\ \rn{STOP\_1}\ (\rn{AND4\_1}\ \rn{STOP\_1}\ \rn{STOP\_1}))\ \ldots
\]}%
of type $\Prf\,((\allI v,\ P_1 \land P_2 \land P_3) \Rightarrow G)$.
Its continuation antecedent $\restm\,(\rho\,v)$ then $\beta$-reduces,
eliminating the result combinators ($\rho$ the derivation above,
$v$ elided):
\begin{align*}
  \restm\,(\rho\,v)
    &≡ \mathsf{conj}\,(\mathsf{res\_ts}\,(\rho\,v)) \\
    &≡ \mathsf{conj}\,(\mathsf{concat}\,(∎ ∷ P_3)\,(\mathsf{concat}\,(∎ ∷ P_2)\,(∎ ∷ P_1))) \\
    &≡ \mathsf{conj}\,(\mathsf{concat}\,(∎ ∷ P_3)\,(∎ ∷ P_2 ∷ P_1)) \\
    &≡ \mathsf{conj}\,(∎ ∷ P_3 ∷ P_2 ∷ P_1) \;≡\; P_3 \land P_2 \land P_1 .
\end{align*}


\paragraph{Arithmetic Side-conditions.}
Some rules invoke \pp's arithmetic solver or equality prover:
\rn{AR4} requires $(E+F) > \mathbf{0}$ for terms \pp{} already believes
positive, and the \rn{ARITH} terminal closes a goal because the
accumulated $\leq$-hypotheses are contradictory.
%
The corresponding lemmas take these facts as explicit $\Prf$-arguments.
\tool{} \emph{generates} each such proof at translation time by running a
small decision procedure over the in-scope hypotheses and assembling the
result from \texttt{B.lp}'s arithmetic lemmas.
%
For \rn{ARITH} that procedure searches for a nonnegative
(Farkas-style~\cite{schrijver1986}) combination of the $\leq \mathbf{0}$
hypotheses that adds up to a false literal.  For example, with
$h_1 : \Prf\,(x \leq \mathbf{0})$ and
$h_2 : \Prf\,(\mathbf{1} - (x + x) \leq \mathbf{0})$ in scope, the combination
$h_2 + 2h_1$ gives $\mathbf{1} - (x + x) + x + x \leq \mathbf{0}$; rewriting
along the generated equality
$e : \Prf\,(\mathbf{1} = (\mathbf{1} - (x + x)) + x + x)$ yields
$\mathbf{1} \leq \mathbf{0}$, a contradiction:
\[
  \mathsf{one\_not\_leq\_zero}\,\bigl(\mathsf{leq\_subst\_l}\ e\
    (\mathsf{add\_leq\_zero}\ \_\ \_\ (\mathsf{add\_leq\_zero}\ \_\ \_\ h_2\ h_1)\ h_1)\bigr)
  : \Prf\,\bot .
\]
%
Permutation proofs are generated likewise whenever \pp{} records a sum in
a different association order than a lemma expects.


\paragraph{Soundness of the rules.}
Every rule symbol is \lpmod{opaque} with a proof body, so \lp{} accepts a
rule only after checking that body against the axiomatic base.  The proofs
follow the structure of the encoding: propositional and quantifier rules
discharge to the classical axioms and the quantifier bridges, conjunction
rules to the $\bwand$ eliminators, equality rules to congruence and
substitution, and arithmetic rules to the ordered-group theory of
\texttt{B.lp}.  No rule is postulated as an axiom; the trusted base is
exactly \texttt{B.lp}'s primitives, the quantifier bridges, and the
$\bwand$/$\Res$ interface.
