% ── eval.tex · §3 ───────────────────────
% Table + prose read numbers from results-data.tex — never hard-code a
% benchmark number in this file.
\section{Implementation and Evaluation}\label{sec:eval}

The core of
\tool{}\footnote{\url{https://github.com/ciaran-matthew-dunne/pp2lp}} % TODO: confirm URL
is an OCaml program of about 4\,400 lines behind a thin command-line
driver; the rule library of \autoref{sec:encoding} is another 4\,300
lines of \lp.
%
A single module holds all rule metadata — arity, argument kinds, and a
\emph{typed} emission strategy per rule — and is the single source of
truth: the emitter dispatches on this variant with an exhaustive match,
so adding a rule without deciding its emission is a compile-time error,
and an unknown rule name in a replay is a runtime error rather than a
silent skip.

\subsection{From a Replay to a \lp{} Proof}\label{sec:pipeline}

% --- pipeline figure ------------------------------------------------
\begin{boxfigure}{fig:pipeline}{%
		The \tool{} pipeline. \pp, instrumented, emits a \trace{} of the
		rules it applied; the \textsc{replay} utility expands this into a
		\replay{} annotating each step with its intermediate goal; \tool{}
		rebuilds the proof tree and emits an \lpf{} tactic script, which
		\lp{} checks against the soundness-proved rule library.}
	\centering
	\alttext{Pipeline diagram with two dashed groups. Left, Atelier B: a
		formula file (.but) maps via PP to a .trace, then via REPLAY to a
		.replay carrying rule and goal lines. An arrow labelled pp2lp maps
		the .replay into the right group, LambdaPi, where the emitted .lp
		file depends on the soundness-proved B.lp and rules/ library; a
		downward arrow labelled lambdapi check yields a tick-or-cross
		verdict.}{%
		\input{fig-pipeline.tex}}
\end{boxfigure}
% --------------------------------------------------------------------

The pipeline (\autoref{fig:pipeline}) has three stages.

\paragraph{Parse.}
A lexer and a Menhir-generated parser read the replay's bracketed rule
lines and formula annotations into a flat list of
(rule, argument, formula) records.

\paragraph{Rebuild the proof tree.}
\tool{} reconstructs the proof tree directly from the replay, consuming
the prefix stream and using each rule's declared arity to attach the
right number of child subproofs.
%
This is where the format's two irregularities (\autoref{sec:pp-syntax})
are handled: phantom normalisation lines are filtered out, and the
result chain that precedes a branching quantifier is reassembled into
that rule's first argument.
%
A mismatch — an arity that does not consume the stream exactly — is
reported with the residual stack.

\paragraph{Emit a tactic script.}
Each obligation becomes one \texttt{opaque symbol} whose statement is
the goal and whose body is a tactic script: each tree node is a
$\refine\;\rn{RULE}\,\ldots$ against the encoded lemma, composed in tree
order (\autoref{fig:reconstruct}).
%
Rules that discharge against a hypothesis receive their
$\Prf$-argument by hypothesis lookup; rules that introduce an antecedent
or a binder push an \texttt{assume}; result chains are emitted as
explicit $\Res$ terms and passed to their branching rule; and solver
side-conditions get freshly generated proof terms
(\autoref{sec:enc-rules}).
%
\lp{} then type-checks the symbol, which closes the loop: \pp's verdict
is now backed by an independently checked proof.
%
The emitted file carries no comments, but the engine records a table
mapping every line to the replay step and \pp{} goal that produced it;
when a check fails, the report shows the offending \lp{} line, the
replay step it came from, and the rule lemma's signature side by side.

% --- reconstruction figure ------------------------------------------
\begin{figure}[t]
  \centering
  \begin{minipage}[c]{0.47\linewidth}\centering
    {\footnotesize\bfseries \pp{} proof replay}\\[3pt]
\begin{codePP}
[OPR1(1)] <delta_pa = TRUE
   => not(a and not(a))>
[AND2] <not(a and not(a))>
[IMP4] <a => not(not(a))>
[NOT2] <not(not(a))>
[AXM3] <a>
\end{codePP}
  \end{minipage}\hfill
  \begin{minipage}[c]{0.49\linewidth}\centering
    {\footnotesize\bfseries emitted \lp{} (verbatim)}\\[3pt]
\begin{codeLP}
opaque symbol zp_noncontra
  (a : Prop) (delta_pa : τ ι) :
  π (delta_pa = BTRUE
     ⇒ ¬ ($\bigwedge$ (∎ ∷ a ∷ ¬ a))) ≔
begin
  assume a delta_pa;
  assume _h1; try rewrite _h1;
  refine AND2 _;
  refine IMP4 _;
  assume _h2;
  refine NOT2 _;
  refine AXM3 _h2
end;
\end{codeLP}
  \end{minipage}
  \caption{Reconstruction of the benchmark \texttt{zp\_noncontra},
    $\lnot(a \land \lnot a)$.
    Each replay line maps to one \refine{} of the corresponding rule
    lemma; the conjunction is the snoc-list $\bwand$ of
    \autoref{sec:enc-formulas}.
    The $\mathtt{delta\_pa} = \mathtt{TRUE}$ premise is a definedness
    flag added by our goal generator; \rn{OPR1} introduces it and
    rewrites with it.}
  \label{fig:reconstruct}
\end{figure}
% --------------------------------------------------------------------

\subsection{Benchmarks}\label{sec:bench}

We exercise \tool{} on three suites (\autoref{tab:results}).
%
\texttt{og} is a fixed set of \ogReplayed{} replays that served as the
initial development corpus.
%
\texttt{prv} is \ab's rule-validation corpus: each goal is an added
proof rule submitted to \pp{} for validation, the prover's primary
industrial role (\autoref{sec:intro}).
%
\texttt{claude} is a synthetic stress and coverage suite of
\claudeGoals{} goals, authored with the help of an LLM-based coding
agent (Claude): it spans every rule family, proof sizes from one to
over 400 replay lines, and deliberately probes shapes that previous
versions of the pipeline mishandled.

\begin{table}[t]
  \centering
  \caption{Results per suite. \emph{Goals} are inputs to the upstream
    \pp/\textsc{replay} toolchain (\texttt{og} has no goal files: its
    replays are checked in); \emph{Replays} are the complete replays
    that toolchain produces (\autoref{sec:limits} accounts for the
    losses); \emph{Verified} are replays \tool{} translates and \lp{}
    type-checks.}
  \label{tab:results}
  \begin{tabular}{l r r r}
    \toprule
    Suite           & Goals        & Replays         & Verified \\
    \midrule
    \texttt{og}     & ---          & \ogReplayed     & \ogPass \\
    \texttt{prv}    & \prvGoals    & \prvReplayed    & \prvPass \\
    \texttt{claude} & \claudeGoals & \claudeReplayed & \claudePass \\
    \midrule
    total           &              & \allReplayed    & \allPass \\
    \bottomrule
  \end{tabular}
\end{table}

Every replay the upstream toolchain produces is reconstructed and
type-checked: \allPass{} proofs, with no failures.
%
\numRulesFired{} of the \numRulesCovered{} encoded rules fire in some
replay.
%
Re-emitting and re-checking the full corpus takes \benchWallAll{} of
wall-clock time on \benchMachine{} (\benchCpuAll{} of \lp{} checking).
%
No emitted proof appeals to the \texttt{trust} oracle — the corpus
contains zero occurrences, enforced by an audit gate — so each checked
obligation rests only on the axiomatic base of \autoref{sec:trust}.

\subsection{Limitations and Discussion}\label{sec:limits}

\paragraph{Upstream replay losses.}
The gap between goals and replays in \autoref{tab:results} is upstream
of \tool: for some provable goals the \textsc{replay} utility omits the
final continuation of a branching quantifier, and for others it aborts
mid-file.
%
Both defects are detected at generation time and the affected goals
dropped loudly.
%
They are not random: set-extensionality and equality-prover-heavy goals
are disproportionately affected — no complete \rn{EGALITE} replay can be
produced from our goal suites, so that emission path is validated
against a hand-completed replay — which means the suites under-test
exactly the rules the equality prover would exercise.

\paragraph{Spec mismatches.}
Encoding every rule against the specification turns the encoding into an
executable test of that document, and it does not always pass: \rn{NRM20}
records its pinned equality at a different conjunct position than
specified, and \rn{OPR1}'s chain form does not follow the result schema
laid out for it.
%
\tool{} implements the observed behaviour; we believe such discrepancies
are themselves a useful by-product for the maintainers of \pp.

\paragraph{Engine partiality.}
A few replay shapes are recognised but rejected: substitution rules
pinning a binder beyond the third slot of a compound binder, and the
chain forms of two arithmetic rules that no producible trace exercises.
%
All of these fail loudly at emission; none silently degrade.

\paragraph{Raw proof obligations.}
We also probed \pp{} on raw industrial obligations from the public
\texttt{apero} dataset~\cite{deharbe2022pog}, translating its POG format
into \pp's rule-validation input.
%
A sampled obligation carries a median of roughly 1\,700 hypothesis
predicates; \pp{} nonetheless proved every convertible obligation with
at most fifty hypotheses, most sampled goals even with all hypotheses
removed, and some thousand-hypothesis obligations given tens of seconds.
%
The bottleneck for certification is again the toolchain around \pp: the
replay utility lost 86 of the 124 traces \pp{} produced on our sample,
and the surviving replays use surface forms no synthetic goal had
reached (strict comparisons, tuple memberships, relational inverse,
sequents inside annotations); 14 of them verify end-to-end today,
including the first machine-produced replays exercising the equality
store.

\paragraph{Use of AI.}
The \texttt{claude} suite was authored by an LLM agent that also helped
develop the tool, so there is a risk of the suite drifting toward what
the pipeline already handles.
%
We mitigate this in two ways: the suite is goal-level only (the agent
cannot influence which rules \pp{} applies), and the real-world
\texttt{og} and \texttt{prv} suites gate every change.

\subsection{Conclusion and Future Work}\label{sec:conclusion}

We have presented \tool, which independently re-checks \ab's Predicate
Prover by reconstructing its proofs in \lp.
%
Unlike approaches that postulate each rule as an axiom, every \pp{} rule
is encoded as a symbol whose body \emph{proves} the rule sound against a
small axiomatic base, and the reconstructed corpus — including \ab's own
rule-validation suite, arithmetic included — type-checks without a
single appeal to an oracle.
%
The most direct future work is upstream: repairing the \textsc{replay}
utility's truncations would recover the dropped goals, in particular the
equality-prover fragment.
%
On the encoding side, the \numRulesTrust{} remaining postulated chain
lemmas can be eliminated by deriving the binder-regrouping isomorphisms
over tuples.
%
Finally, this work folds \pp{} into the \dedukti{} interoperability
ecosystem: a natural next step within the ICSPA project is exporting
checked \ab{} proofs to set-theory-based assistants such as TLA$^+$'s
TLAPS, Isabelle/ZF, and Megalodon.

\subsubsection*{Acknowledgements.}
This work is supported by the ANR project ICSPA
(ANR-21-CE25-0015) in collaboration with CLEARSY.
% TODO: confirm grant number and any additional acknowledgements.
