% ── intro.tex · §1 ──────────────────────────────
% other kinds of proof obligations. well-definedness, overflow, constants.
%
\section{Introduction}\label{sec:intro}

The \bmethod~\cite{abrial1996bbook} is a formal method for developing software
by \emph{stepwise refinement}: a development starts from an abstract
machine (i.e., state, invariants, and operations expressed in first-order
logic and set theory) and proceeds through proved refinement steps down
to executable code.
%
It has a long industrial record in the verification of safety-critical systems,
most prominently in rail transport: the automatic train control of Paris métro
line~14 was developed in B~\cite{behm1999meteor}.
%
\ab~\cite{clearsy_atelierb} is a software package supporting the method,
developed and distributed by CLEARSY.
%
% well-definedness conditions.
For every machine and refinement step, it generates \emph{proof obligations}
(e.g., sequents whose validity establishes that an invariant
       is preserved or a refinement is correct)
and discharges them with automated provers, leaving the remainder to
interactive proof.

\ab's main prover applies rules drawn from a large rule
base~\cite{lecomte2024provingb}, and users routinely extend this base
with bespoke rules capturing the reasoning of their domain; an added
rule must itself be validated.
%
The \emph{Predicate Prover} (\pp) is \ab's automated theorem prover,
and serves both roles: it discharges proof obligations directly, and it
is the standard tool for validating added rules.
%
Despite B's set-theoretic surface, \pp{} is almost entirely a prover for
first-order logic with linear arithmetic: a pre-processing pass reduces
the set theory of an obligation to first-order formulas over membership,
and the set operators that remain reach \pp{} as uninterpreted function
symbols (\autoref{sec:pp-syntax}).
%
\pp{} is closed-source: its verdicts are trusted by \ab{} without
independent verification.
%
This is in contrast to setups like Isabelle's
Sledgehammer~\cite{sledgehammer}, where proofs found by external provers
are reconstructed within the proof assistant's kernel rather than taken
on faith.
%
We address this trust gap with \tool, a tool that reconstructs \pp{} proofs in
\lp~\cite{hondet:hal-02981561}, a proof assistant based on the
\lpcmr~\cite{CousineauDowek}.
%
We encode each \pp{} inference rule as a \lp{} symbol whose type captures
the rule's premises and conclusion, and whose body is a proof that
demonstrates the correctness of the rule with respect to classical logic
and basic set theory.
% and arithmetic?
% some PP inference rules just correspond to tactics.
%

% 'achieve this' implies
Fortunately, \pp{} is instrumented to record a compact \emph{trace} listing
the inference rules it applied during a proof; another utility expands this
trace into a \emph{replay} that annotates each step with the rule applied and
the intermediate goal it was applied to.
%
Given a replay, \tool{} parses it, builds a proof tree, and emits a \lp{}
tactic script that composes the encoded rules following the tree, yielding
a complete, independently checkable proof.
%
We evaluate our tool on benchmark suites containing rule validations
from real-world \ab{} use and a large synthetic suite of goals generated
by Claude Opus 4.8.
%
Every proof accepted by the replay utility (\allReplayed{} in total) is
reconstructed and type-checked, using no unproved assumption beyond a small,
fixed axiomatic base.
%
% We additionally probe \tool{} on raw proof obligations from real-world
% \ab{} developments, where the chief obstacle turns out to be upstream
% replay loss rather than reconstruction (\autoref{sec:limits}).

\paragraph{Related Work.}
This work is carried out within the ANR project ICSPA (\emph{Interoperable
and Confident Set-based Proof Assistants})\footnote{\url{https://icspa.inria.fr}},
which aims to connect the set-theoretic proof systems of industrial formal
methods (B and TLA$^+$) to the \dedukti{} ecosystem for independent verification
and interoperability.
%
Sibling efforts in the project encode Event-B's mathematical language
and its Rodin-produced proofs in \lp~\cite{grieu2024eventb,
grieu2025jfla}, and translate Event-B developments to
TLA$^+$~\cite{grieu2025eventbtla}.
% TODO(authors): Catherine's AFADL 2026 companion paper (Lille, 2--3
% June 2026) has no HAL record yet — add the citation once the actes
% are deposited (nothing to cite as of 2026-06-12).
The work presented here complements these approaches, since \pp{} is
used in Rodin.

% TODO()
The reduction of proof checking to type checking originates with the
Edinburgh Logical Framework~\cite{harper1993lf}, whose underlying
calculus is the dependently-typed $\lambda\Pi$-calculus.
%
\dedukti{} extends this calculus with user-declared rewrite rules,
yielding the \lpcmr~\cite{CousineauDowek, assaf2023dedukti}; \lp{} is
its interactive front end~\cite{hondet:hal-02981561}.
%
Replaying automated provers inside such checkers is an active line of
work: SMT proofs have been reconstructed in
\lp~\cite{coltellacci2026acta} — including proofs of TLAPS obligations,
TLA$^+$'s proof system~\cite{coltellacci2023abz} — and in
Lean~\cite{mohamed25leansmt}.
% LEO-III, Zenon Modulo, why3, hammers
%GB: some words on hammer-like tools?

Confidence in \ab's proof machinery has been pursued from several
directions.
%
% strategies != rewrite rules.
Cirstea and Kirchner formalised \pp's inference rules as rewriting
strategies in ELAN~\cite{cirstea1998elan}; their encoding re-implements
the prover's first-order core, but covers neither arithmetic nor any
soundness argument for the rules themselves.
%
Jaeger and Dubois~\cite{jaeger2007trustb} asked the trust question
directly, giving a deep embedding of B's logic in Rocq (BiCoq);
the BCARe framework of Jacquel et~al.~\cite{jacquel2011bcare, jacquel2015sosym}
works over a similar embedding to verify the rules added by users
to \ab's rule base.
%
None of these efforts checks the deductions that \pp{}
actually performs on a given obligation.
%
Our work targets exactly that gap: each \pp{} run is re-checked
against rule lemmas that are themselves proved sound.
%
% Trelat BeeR.

%
The BWare project~\cite{bware2014} took a different route to the same
trust concern: it discharged B proof obligations with external provers
--- notably Zenon Modulo, whose proofs are checked by
\dedukti~\cite{bury2015bset} --- thereby replacing \pp{} rather than
formally verifying its output.
% Terlat Barrel
In the same spirit, Trélat~\cite{trelat2025smt} gives a provably sound
encoding of B proof obligations into SMT-LIB, as does
Defourné~\cite{defourne2025scp} for TLA$^+$ obligations.
%
Closest to our setting, Stolze et~al.~\cite{stolze2024dedukti} export
\ab{} proof \emph{obligations} to \dedukti{} and report preliminary
experiments towards sharing \ab{} proofs across the \dedukti{} ecosystem.
%
% TODO(authors): confirm with Hermant/D\'eharbe who built the trace
% instrumentation and REPLAY before claiming any tooling lineage from
% stolze2024dedukti in print.  Public evidence points to Atelier B
% itself (REPLAY ships as an Atelier B kernel image, and the trace
% format is documented in the Atelier B manual pp-trace-manref.pdf);
% the Stolze et al. preprint mentions neither traces nor replays.

\paragraph{Outline.}
\autoref{sec:encoding} recalls \pp's language and proof format, and
presents our encoding of its rules in \lp.
%
\autoref{sec:eval} describes the translation pipeline and its
implementation, reports benchmark results, and discusses limitations
and future work.
