% ── preliminaries.tex · §2 Preliminaries (PP & LP) ─────────────────
\section{Preliminaries}\label{sec:prelim}

\subsection{The Predicate Prover}\label{sec:pp-syntax}

The syntax for the \emph{predicates} and \emph{expressions} of \pp{} is
given as follows:
\begin{align*}
	E, F & \Coloneqq x \mid n \mid E \mapsto F \mid F(E_1,\ldots,E_k)
	\mid E + F \mid E - F \mid -E                                     \\
	P, Q & \Coloneqq E \mid E = F \mid E \leq F
	\mid E_1,\ldots,E_k \in F
	\mid \lnot P \mid P \land Q \mid P \lor Q \mid P \Rightarrow Q
	\mid P \Leftrightarrow Q
	\mid \mathcal{Q}\,\vec{x} \cdot P
\end{align*}
where $x$ ranges over identifiers, $n$ over numerals, $\mapsto$ is B's pairing,
$F(E_1,\ldots,E_k)$ applies $F$ to arguments, and membership relates a tuple of
expressions to an expression.
%
An expression may itself stand as an atomic predicate, shown by the leading
$E$ in the production for $P$.
%
A quantifier $\mathcal{Q} \in \{\forall, \exists, \Dia, \Hrt\}$ binds a
list of variables; $\forall (x,y) \cdot P$ is distinct from $\forall x \cdot \forall y
	\cdot P$, and dedicated rules (\rn{ALL1--4}, \rn{XST1--4}) regroup the
latter into the former during proof search.
%
The quantifiers $\Dia$ and $\Hrt$ are universal quantifiers idiosyncratic
to \pp's normalisation procedure; semantically both are $\forall$, with
$\Dia$ marking a normalised universal on the goal side and $\Hrt$ one on the
hypothesis side.

Although B is a set-theoretic formalism, \pp{} proves first-order logic with
linear arithmetic: pre-processing reduces much of an obligation's set theory to
formulas over membership before \pp{} sees it.
%
The set-theoretic operators that survive this reduction (intervals, maplets,
and the set and relational operators of \autoref{app:operators}) reach
\pp{} as uninterpreted function symbols: \pp{} has no set-theory rules, and
such terms are only ever copied, compared for equality, or instantiated.

\pp{} proves \emph{sequents} $H \seq P$, where $H$ is a list of
hypothesis predicates and $P$ a goal predicate.
%
\pp's inference rules are specified in the annex of its internal
documentation~\cite{pp_spec}.  There are roughly \numRulesSpec{} rules,
grouped into families: propositional (\rn{AND}, \rn{OR}, \rn{IMP}, \rn{EQV},
\rn{NOT}), hypothesis discharge (\rn{AXM1--9}), quantifiers
(\rn{ALL1--9}, \rn{XST1--8}), equality
(\rn{EVR}, \rn{EAXM}, \rn{EQC}, \rn{EQS}, \rn{ECTR}, \ldots), booleans
(\rn{BOOL}), linear arithmetic (\rn{AR1--13}),
and normalisation (\rn{NRM1--30}), among others.
%
Each rule reduces the current goal to zero, one, or two
premises, possibly under computational side-conditions
(``$P$ occurs in the list of hypotheses $H$'',
``$a$ and $b$ are integer literals such that $a + b < 0$'', etc.).

Alongside sequents, \pp's proofs manipulate a second judgment: we write
$(H \seq P) \yields [Q_1, \ldots, Q_n]$ for ``under $H$, $P$ normalises to
the list of predicates $[Q_1, \ldots, Q_n]$''.  This list is a
\emph{result}, whose conjunction is the normal form of $P$.
%
Most classical rules come paired with a result form (written with a
\texttt{\_1} suffix): a rule whose conclusion is a result.  A result rule is
given by one of three schemas: the empty list (the formula is discharged),
a singleton $[Q]$ (a single transformed formula, produced by a leaf such as
\rn{STOP\_1}), or the concatenation of its two premises' lists (a combination).
%
For instance, \rn{AND4} introduces a conjunct, and its result form
\rn{AND4\_1} (the combination schema) concatenates the two premises'
lists:
\begin{center}
  \begin{prooftree}
    \hypo{H \seq Q} \hypo{H \seq P}
    \infer2[\rn{AND4}]{H \seq P \land Q}
  \end{prooftree}
  \qquad
  \begin{prooftree}
    \hypo{(H \seq Q) \yields R_1} \hypo{(H \seq P) \yields R_2}
    \infer2[\rn{AND4\_1}]{(H \seq P \land Q) \yields R_1 \app R_2}
  \end{prooftree}
\end{center}

Only \rn{ALL7} and \rn{XST8} consume a result, each turning a
result derivation back into an ordinary sequent subproof.
%
\rn{ALL7} proves $(\forall x \cdot P) \Rightarrow G$ by normalising $P$'s body
pointwise to a list $R_1,\ldots,R_n$ and continuing under
$(\Dia x \cdot R_1 \land \ldots \land R_n) \Rightarrow G$, beneath the
post-normalisation universal $\Dia$.
%
\rn{XST8} is its existential dual: to prove $\exists x \cdot P$ it normalises
$\lnot P$ to a list $R_1,\ldots,R_n$ and refutes their conjunction closed
under the same post-normalisation universal $\Dia$ --- the same shape as
\rn{ALL7}, with $\bot$ in place of the goal $G$.

\pp{} records a \trace{} of the rules it applied; a companion utility
re-runs the proof and expands the trace into a \replay{} that annotates each
step with the goal it was applied to (\autoref{fig:replay}).
%
Each replay line carries a rule name and, in parentheses, an optional
rule-specific argument (for instance, an instantiation index).
%
A rule's line precedes the lines of its subproofs, with two exceptions.
%
First, the result derivation feeding \rn{ALL7} or \rn{XST8} appears
before the rule that consumes it.
%
Second, the replay interleaves marker lines that do not correspond to
inference rules: \texttt{NRM} and \texttt{STOP\_NORM}, which bracket
\pp's normalisation passes, and \texttt{FIN}, which carries the predicate
such a pass produced.
%
\tool{} filters these out while rebuilding the tree.
%

\subsection{\lp}\label{sec:lp}

\lp~\cite{hondet:hal-02981561} is a proof assistant for the
\lpcmr~\cite{CousineauDowek}, the calculus underlying the \dedukti{}
logical framework~\cite{assaf2023dedukti}.
%
Unlike \dedukti{}, a bare type-checker for the same calculus, \lp{}
adds implicit arguments $[x:A]$ and an interactive tactic language.
%
The term syntax is given as follows:
\[
	t, u \Coloneqq \TYPE \mid \KIND \mid x \mid c
	\mid t\,u \mid \lambda (x : t),\, u \mid \Pi (x : t),\, u
\]
where $x$ ranges over variables, $c$ ranges over declared constants,
and terms are identified up to $\alpha$-equivalence.
%
Typing is standard for the \lpcmr{} (the full system is in
\autoref{app:typing}); its \emph{definitional equality} (i.e., $A \Equiv B$)
is generated by $\beta$-equivalence and the user-declared rewrite rules.
%
A \lp{} development is a sequence of declarations:
\[
  \begin{array}{rcl}
    d &\Coloneqq& m^{*}\;\text{\lpkw{symbol}}\;x\;p^{*} : t\;[\,≔ b\,]
        \;\mid\; \text{\lpkw{rule}}\;\ell ↪ r
        \;\mid\; \text{\lpkw{inductive}}\;x\;p^{*} : t ≔ \mathit{cons} \\
    p &\Coloneqq& (x : t) \;\mid\; [x : t] \\
    m &\Coloneqq& \text{\lpmod{constant}} \;\mid\; \text{\lpmod{injective}} \;\mid\; \text{\lpmod{opaque}} \\
    b &\Coloneqq& t \;\mid\; \text{\lpkw{begin}}\;\mathit{tac}^{*}\;\text{\lpkw{end}}
  \end{array}
\]
The \lpkw{symbol} command introduces a typed constant with explicit
$(x : A)$ and implicit $[x : A]$ parameters $p$ and an optional definition
$b$; \lpkw{rule} adds a rewrite rule over pattern variables \texttt{\$x},
checked for type-preservation at declaration time; \lpkw{inductive}
generates an inductive type with constructors $\mathit{cons}$ and a recursor.
%
Modifiers $m$ refine a symbol's status: \lpmod{constant} symbols admit no
rewrite rule at their head, \lpmod{injective} symbols may be inverted
during unification, and the definition of an \lpmod{opaque} symbol is
sealed once type-checked.
%
A definition $b$ is a term or a proof script (\lpkw{begin} \ldots{}
\lpkw{end}) whose tactics (e.g., \lpkw{assume}, \lpkw{refine},
\lpkw{rewrite}, \lpkw{have}, \lpkw{induction}) elaborate to a proof term
that the kernel re-checks.  The grammars above are abridged to the constructs
the body uses; \autoref{app:syntax} gives the full grammar of our \lp{}
fragment, with its inductive and tactic syntax.

We build on \lp's standard library~\cite{Blanqui2021}: object-level types are
encoded in a Tarski-style universe $\Setu$ with $\El : \Setu \to \TYPE$
mapping each set code to its type of inhabitants, and a type $\Prop$ of
propositions with $\Prf : \Prop \to \TYPE$ mapping each proposition
to its type of proofs.
%
It also supplies the booleans $\mathbb{B}$, integers $\mathbb{Z}$, and
naturals $\mathbb{N}$ (with integer literals and arithmetic), which our
encoding builds on (\autoref{sec:enc-formulas}).

